Why shouldn't you use numerical tests to diagnose the linear regression models?

### 2022-07-07 paper_comparison/paper_comparison/paper_comparison.Rmd
update intro

---
title: |
  Why shouldn't you use numerical tests to diagnose the linear regression models?
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
bibliography: paper.bib
abstract: |
  Abstract to fill.
keywords: |
  visual inference; model diagnostics;
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

@@

@@

As discussed in @cook1982residuals, most residual-based tests for a particular type of departures from model assumptions are sensitive to other types of departures. It is likely the null hypothesis is correctly rejected but for the wrong reason, which is also known as the "Type III error". Additionally, outliers will often incorrectly trigger the rejection of null hypothesis despite the residuals are well-behaved [@cook_applied_1999]. This can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers. @montgomery_introduction_2012 stated that based on their experience, statistical tests are not widely used in regression diagnostics. The same or even larger amount of information can be provided by diagnostic plots than the corresponding tests in most empirical studies. Not to mention, it is almost impossible to have an exactly correctly specified model in reality. There is a well-known aphorism in statistic stated by George Box - "All models are wrong, but some are useful". This indicates proper hypothesis tests will always reject the null hypothesis as long as the sample size is large enough. The outcome "Not reject" can be interpreted as either "effect size is small" or "sample size is small". The outcome "reject" still doesn't inform us whether and how much the model defects are of actual consequence to the inference and prediction. But still, the effectiveness of statistical tests shall not be disrespected. Statistical tests have chance to provide analysts with unique information. There are situations where no suitable diagnostic plots can be found for a particular violation of the assumptions, or excessive diagnostic plots need to be checked. One will have no choice but to rely on statistical tests if there is any. A good regression diagnostic practice should be a balanced combination of both methods.

Diagnostic plots are

<!-- However, unlike confirmatory data analysis built upon rigorous statistical procedures, e.g., hypothesis testing, visual diagnostics relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding an over-interpretation of the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015]. -->

<!-- Recently, a new branch of statical inference  -->

<!-- Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots [@gelman_exploratory_2004]. Later, in the Bayesian context, data plots was systematically considered as model diagnostics by taking advantage of the data simulated from the assumed statistical models [@gelman_bayesian_2003; @gelman_exploratory_2004].  -->

>>>>

---
title: |
  Why shouldn't you use numerical tests to diagnose the linear regression models?
type: ARTICLE TEMPLATE
author:
  - name: Weihao Li
    affil: a
    email: weihao.li@monash.edu
  - name: Dianne Cook
    affil: a
    email: dicook@monash.edu
  - name: Emi Tanaka
    affil: a
    email: emi.tanaka@monash.edu
affiliation:
  - num: a
    address: |
      Department of Econometrics and Business Statistics, Monash University, Clayton, VIC, Australia
bibliography: paper.bib
abstract: |
  Abstract to fill.
keywords: |
  visual inference; regression diagnostics;
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

@@

## Diagnostic plots

@@

## Hypothesis testing

@@

As discussed in @cook1982residuals, most residual-based tests for a particular type of departure from model assumptions are sensitive to other types of departures. It is likely the null hypothesis is correctly rejected but for the wrong reason, which is also known as the "Type III error". Additionally, outliers will often incorrectly trigger the rejection of the null hypothesis despite the residuals being well-behaved [@cook_applied_1999]. This can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers. @montgomery_introduction_2012 stated that based on their experience, statistical tests are not widely used in regression diagnostics. The same or even larger amount of information can be provided by diagnostic plots than the corresponding tests in most empirical studies. Not to mention, it is almost impossible to have an exactly correctly specified model in reality. There is a well-known aphorism in statistics stated by George Box - "All models are wrong, but some are useful". This indicates proper hypothesis tests will always reject the null hypothesis as long as the sample size is large enough. The outcome "Not reject" can be interpreted as either "effect size is small" or "sample size is small". The outcome "reject" still doesn't inform us whether and how much the model defects are of actual consequence to the inference and prediction. But still, the effectiveness of statistical tests shall not be disrespected. Statistical tests have a chance to provide analysts with unique information. There are situations where no suitable diagnostic plots can be found for a particular violation of the assumptions, or excessive diagnostic plots need to be checked. One will have no choice but to rely on statistical tests if there are any. A good regression diagnostic practice should be a balanced combination of both methods.

## Visual inference

However, unlike hypothesis testing built upon rigorous statistical procedures, reading diagnostic plots relies on graphical perception - human’s ability to interpret and decode the information embedded in the graph [@cleveland_graphical_1984], which is to some extent subjective. Further, visual discovery suffers from its unsecured and unconfirmed nature where the degree of the presence of the visual features typically can not be measured quantitatively and objectively, which may lead to over or under-interpretations of the data. One such example is finding an over-interpretation of the separation between gene groups in a two-dimensional projection from a linear discriminant analysis when in fact there are no differences in the expression levels between the gene groups and separation is not an uncommon occurrence [@roy_chowdhury_using_2015].

Visual inference was first introduced in a 1999 Joint Statistical Meetings (JSM) talk with the title "Inference for Data Visualization" by @buja_inference_1999 as an idea to address the issue of valid inference for visual discoveries of data plots [@gelman_exploratory_2004]. Later, in the Bayesian context, data plots was systematically considered as model diagnostics by taking advantage of the data simulated from the assumed statistical models [@gelman_bayesian_2003; @gelman_exploratory_2004].

It was surprising that the essential components of visual inference had actually been established in @buja_inference_1999, but it was not until 10 years later that @buja_statistical_2009 formalized it as an inferential framework to extend confirmatory statistics to visual discoveries. This framework redefines the test statistics, tests, null distribution, significance levels and $p$-value for visual discovery modelled on the confirmatory statistical testing. Figure \@ref(fig:parallelism) outlines the parallelism between conventional tests and visual discovery.

![(ref:parallelism) \label{fig:parallelism}](figures/rsta2009012001.jpg){width=450 height=341}

(ref:parallelism) Parallelism between multiple quantitative testing and visual discovery [@buja_statistical_2009]. Visible features in a plot are viewed as a collection of test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$, and any visual discoveries that are inconsistent with the null hypothesis are treated as evidence against the null. For regression diagnostics, the null hypothesis would be the assumed model, and visual discoveries could be any visual features in favour of any alternatives.

In visual inference, a collection of test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)$ is defined, where $\boldsymbol{\mathrm{y}}$ is the data and $I$ is a set of all possible visual features. @buja_statistical_2009 described each of the test statistics $T^{(i)}(\boldsymbol{\mathrm{y}})$ as a measurement of the degree of presence of a visual feature. Alternatively, @majumder_validation_2013 avoids the use of visual features and defined the visual statistics $T(.)$ as a mapping from a dataset to a data plot. Both definitions of visual test statistics are valid, but in the rest of the report the first definition will be used as it covers some details needed by the following discussion. A visual discovery is defined as a rejection of a null hypothesis, and the same null hypothesis can be rejected by many different visual discoveries [@buja_statistical_2009]. For regression diagnostics, the null hypothesis would be the assumed model, while the visual discoveries would be any findings that are inconsistent with the null hypothesis. The same regression model can be rejected by many reasons with residual plot, including non-linearity and heteroskedasticity as shown in Figure \@ref(fig:residual-plot-cubic-heter).

```{r residual-plot-cubic-heter, fig.cap='Residuals vs. fitted values plot for a classical linear regression model. The residuals are produced by fitting a two-predictor multiple linear regression model with data generated from a cubic linear model. From the residual plot, "butterfly shape" can be observed which generally would be interpretd as evidence of heteroskedasticity. Further, from the outline of the shape, nonlinear patterns exist. Both visual discoveries are evidence against the null hypothesis, though heteroskedasticity actually does not exist in the data generating process.'}

cm <- cubic_model(a = 0, b = 100, c = 1)

cm$gen(1000, fit_model = TRUE) %>%
cm$plot() +
  theme_light()
```

### Sampling from the null distribution {#se:sampling-from-null}

The null distribution of plots refers to the infinite collection of plots of null datasets sampled from $H_0$. It is defined as the analogue of the null distribution of test statistics in conventional test [@buja_statistical_2009]. In practice, a finite number of plots of null datasets could be generated, called null plots. 
In the context of regression diagnostics, sampling data from $H_0$ is equivalent to sampling data from the assumed model. As @buja_statistical_2009 suggested, $H_0$ is usually composited by a collection of distributions controlled by nuisance parameters.  Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling.

The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja_statistical_2009]. Suppose there exists a minimal sufficient statistic $\boldsymbol{S}(\boldsymbol{y})$ under the null hypothesis, any null datasets $\boldsymbol{y^{*}}$ should fulfil the condition $\boldsymbol{S}(\boldsymbol{y}) = \boldsymbol{s}$. Using the classical normal linear regression model as example, the minimal sufficient statistic is $\boldsymbol{S}(\boldsymbol{y}) = (\hat{\boldsymbol{\beta}}, \boldsymbol{e}'\boldsymbol{e})$, where $\hat{\boldsymbol{\beta}}$ are the coefficient estimators and $\boldsymbol{e}'\boldsymbol{e}$ is the residual sum of square. Alternatively, the minimal sufficient statistic can be constructed as $\boldsymbol{S}(\boldsymbol{y}) = (\hat{\boldsymbol{y}}, ||\boldsymbol{e}||)$, where $\hat{\boldsymbol{y}}$ are the fitted values and $||\boldsymbol{e}||$ is the length of residuals, which is more intuitive as suggested by @buja_statistical_2009. Since the fitted values are held fixed, the variation can only occur in the residual space. And because the length of residual is also held fixed, residuals obtained from a null dataset has to be a random rotation of $\boldsymbol{e}$ in the residual space. With this property, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the regressors, then rescaling it by the ratio of residual sum of square in two regressions.

### Lineup protocol {#se:lineup}

With the simulation of null plots being provided, another aspect of hypothesis testing that needs to be addressed is the control of false positive rate or Type I error. Any visual statistic $T^{(i)}(\boldsymbol{\mathrm{y}})$ needs to pair with a critical value $c^{(i)}$ to form a hypothesis test. When a visual feature $i$ is discovered by the observer from a plot, the corresponding visual statistic $T^{(i)}(\boldsymbol{\mathrm{y}})$ may not be known as there is no general agreement on the measurement of the degree of presence of a visual feature. It is only the event that $T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}$ is confirmed. Similarly, if any visual discovery is found by the observer, we say, there exists $i \in I:~T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}$ [@buja_statistical_2009].

Using the above definition, the family-wise Type I error can be controlled if one can provide the collection of critical values $c^{(i)}~(i \in I)$ such that $P(\mathrm{there~exists~} i \in I: T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}|\boldsymbol{\mathrm{y}}) \leq \alpha$, where $\alpha$ is the significance level. However, since the quantity of $T^{(i)}(\boldsymbol{\mathrm{y}})$ may not be known, such collection of critical values can not be provided.

@buja_statistical_2009 proposed the lineup protocol as a visual test to calibrate the Type I error issue without the specification of $c^{(i)}~(i \in I)$. It is inspired by the "police lineup" or "identity parade" which is the act of asking the eyewitness to identify criminal suspect from a group of irrelevant people. The protocol consists of $m$ randomly placed plots, where one plot is the actual data plot, and the remaining $m - 1$ plots have the identical graphical production as the data plot except the data has been replaced with data consistent with the null hypothesis. Then, an observer who have not seen the actual data plot will be asked to point out the most different plot from the lineup.

Under the null hypothesis, it is expected that the actual data plot would have no distinguishable difference with the null plots, and the probability of the observer correctly picks the actual data plot is $1/m$. If we reject the null hypothesis as the observer correctly picks the actual data plot, then the Type I error of this test is $1/m$.

This provides us with an mechanism to control the Type I error, because $m$ - the number of plots in a lineup can be chosen. A larger value of $m$ will result in a smaller Type I error, but the limit to the value of $m$ depends on the number of plots a human is willing to view [@buja_statistical_2009]. Typically, $m$ will be set to $20$ which is equivalent to set $\alpha = 0.05$, a general choice of significance level for conventional testing among statisticians.

Further, if we involve $K$ independent observers in a visual test, and let $X$ be a random variable denoting the number of observers correctly picking the actual data plot. Then, under the null hypothesis $X \sim \mathrm{Binom}_{K,1/m}$, and therefore, the $p$-value of a lineup of size $m$ evaluated by $K$ observer is given as
\begin{equation} \label{eq:pvaluesingle}
P(X \geq x) = \sum_{i=x}^{K}{{K}\choose{i}}\left(\frac{1}{m}\right)^i\left(\frac{m-1}{m}\right)^{k-i},
\end{equation}

where $x$ is the realization of number of observers correctly picking the actual data plot [@majumder_validation_2013].

The multiple individuals approach avoids the limit of $m$, while provides visual tests with $p$-value much smaller than $0.05$. In fact, the lower bound of $p$-value decreases exponentially as $K$ increases. With just $4$ individuals and $20$ data plots in a lineup, the $p$-value could be as small as $0.0001$. Additionally, by involving multiple observers, variation of individual ability to read plots can be addressed to some degree as different opinions about visual discoveries can be collected.

Compared to the conventional test, whose power only depends on the parameter of interest $\theta$, several studies [see @hofmann_graphical_2012; @majumder_validation_2013; @majumder_human_2014; @roy_chowdhury_using_2015; @loy_variations_2016] have shown the power of the visual test is subject-specific. Thus, to be able to account for individual's ability, an individual is required to evaluate multiple lineups [@majumder_validation_2013].

Suppose individuals have the same ability and a lineup has been evaluated by multiple individuals, under the alternative hypothesis, the estimated power for a lineup can be expressed as $\hat{p} = x/K$, the estimated probability of identifying the actual data plot from the lineup. If the individual skill needs to be taken into account, and $L$ lineups have been evaluated by $K$ individuals, @majumder_validation_2013 suggests that mixed effects logistic regression model can be fit as:

$$g(p_{li}) = W_{li}\delta + Z_{li}\tau_{li},$$
where $g(.)$ is the logit link function $g(p) = log(p)  - log(1-p)$; $0 \leq p \leq 1$. $W_{li}$, $1 \leq i \leq K$, $1 \leq l \leq L$, is the covariate matrix including lineup-specific elements and demographic information of individuals, and $\delta$ is a vector of parameters. $Z$ is the random effects matrix, and $\tau$ is a vector of variables follow $N(\boldsymbol{0},\sigma_{\tau}\boldsymbol{I}_{KL\times KL})$.

Then, the estimated power for lineup $l$ and individual $i$ can be calculated as $\hat{p}_{li} = g^{-1}(W_{li}\hat{\delta} + Z_{li}\hat{\tau}_{li})$ [@majumder_validation_2013].

### 2022-07-07 paper_comparison/paper_comparison/paper_comparison.tex
update intro

\begin{keywords}
visual inference; model diagnostics;
\end{keywords}

@@

@@

@@

As discussed in \citet{cook1982residuals}, most residual-based tests for
a particular type of departures from model assumptions are sensitive to
other types of departures. It is likely the null hypothesis is correctly
rejected but for the wrong reason, which is also known as the ``Type III
error''. Additionally, outliers will often incorrectly trigger the
rejection of null hypothesis despite the residuals are well-behaved
\citep{cook_applied_1999}. This can be largely avoided in diagnostic
plots as experienced analysts can evaluate the acceptability of
assumptions flexibly, even in the presence of outliers.
\citet{montgomery_introduction_2012} stated that based on their
experience, statistical tests are not widely used in regression
diagnostics. The same or even larger amount of information can be
provided by diagnostic plots than the corresponding tests in most
empirical studies. Not to mention, it is almost impossible to have an
exactly correctly specified model in reality. There is a well-known
aphorism in statistic stated by George Box - ``All models are wrong, but
some are useful''. This indicates proper hypothesis tests will always
reject the null hypothesis as long as the sample size is large enough.
The outcome ``Not reject'' can be interpreted as either ``effect size is
small'' or ``sample size is small''. The outcome ``reject'' still
doesn't inform us whether and how much the model defects are of actual
consequence to the inference and prediction. But still, the
effectiveness of statistical tests shall not be disrespected.
Statistical tests have chance to provide analysts with unique
information. There are situations where no suitable diagnostic plots can
be found for a particular violation of the assumptions, or excessive
diagnostic plots need to be checked. One will have no choice but to rely
on statistical tests if there is any. A good regression diagnostic
practice should be a balanced combination of both methods.

Diagnostic plots are

>>>>

\begin{keywords}
visual inference; regression diagnostics;
\end{keywords}

@@

\hypertarget{diagnostic-plots}{%
\subsection{Diagnostic plots}\label{diagnostic-plots}}

@@

\hypertarget{hypothesis-testing}{%
\subsection{Hypothesis testing}\label{hypothesis-testing}}

@@

As discussed in \citet{cook1982residuals}, most residual-based tests for
a particular type of departure from model assumptions are sensitive to
other types of departures. It is likely the null hypothesis is correctly
rejected but for the wrong reason, which is also known as the ``Type III
error''. Additionally, outliers will often incorrectly trigger the
rejection of the null hypothesis despite the residuals being
well-behaved \citep{cook_applied_1999}. This can be largely avoided in
diagnostic plots as experienced analysts can evaluate the acceptability
of assumptions flexibly, even in the presence of outliers.
\citet{montgomery_introduction_2012} stated that based on their
experience, statistical tests are not widely used in regression
diagnostics. The same or even larger amount of information can be
provided by diagnostic plots than the corresponding tests in most
empirical studies. Not to mention, it is almost impossible to have an
exactly correctly specified model in reality. There is a well-known
aphorism in statistics stated by George Box - ``All models are wrong,
but some are useful''. This indicates proper hypothesis tests will
always reject the null hypothesis as long as the sample size is large
enough. The outcome ``Not reject'' can be interpreted as either ``effect
size is small'' or ``sample size is small''. The outcome ``reject''
still doesn't inform us whether and how much the model defects are of
actual consequence to the inference and prediction. But still, the
effectiveness of statistical tests shall not be disrespected.
Statistical tests have a chance to provide analysts with unique
information. There are situations where no suitable diagnostic plots can
be found for a particular violation of the assumptions, or excessive
diagnostic plots need to be checked. One will have no choice but to rely
on statistical tests if there are any. A good regression diagnostic
practice should be a balanced combination of both methods.

\hypertarget{visual-inference}{%
\subsection{Visual inference}\label{visual-inference}}

However, unlike hypothesis testing built upon rigorous statistical
procedures, reading diagnostic plots relies on graphical perception -
human's ability to interpret and decode the information embedded in the
graph \citep{cleveland_graphical_1984}, which is to some extent
subjective. Further, visual discovery suffers from its unsecured and
unconfirmed nature where the degree of the presence of the visual
features typically can not be measured quantitatively and objectively,
which may lead to over or under-interpretations of the data. One such
example is finding an over-interpretation of the separation between gene
groups in a two-dimensional projection from a linear discriminant
analysis when in fact there are no differences in the expression levels
between the gene groups and separation is not an uncommon occurrence
\citep{roy_chowdhury_using_2015}.

Visual inference was first introduced in a 1999 Joint Statistical
Meetings (JSM) talk with the title ``Inference for Data Visualization''
by \citet{buja_inference_1999} as an idea to address the issue of valid
inference for visual discoveries of data plots
\citep{gelman_exploratory_2004}. Later, in the Bayesian context, data
plots was systematically considered as model diagnostics by taking
advantage of the data simulated from the assumed statistical models
\citep{gelman_bayesian_2003, gelman_exploratory_2004}.

It was surprising that the essential components of visual inference had
actually been established in \citet{buja_inference_1999}, but it was not
until 10 years later that \citet{buja_statistical_2009} formalized it as
an inferential framework to extend confirmatory statistics to visual
discoveries. This framework redefines the test statistics, tests, null
distribution, significance levels and \(p\)-value for visual discovery
modelled on the confirmatory statistical testing. Figure
@ref(fig:parallelism) outlines the parallelism between conventional
tests and visual discovery.

\begin{figure}
\centering
\includegraphics[width=4.6875in,height=3.55208in]{figures/rsta2009012001.jpg}
\caption{(ref:parallelism) \label{fig:parallelism}}
\end{figure}

(ref:parallelism) Parallelism between multiple quantitative testing and
visual discovery \citep{buja_statistical_2009}. Visible features in a
plot are viewed as a collection of test statistics
\(T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)\), and any visual
discoveries that are inconsistent with the null hypothesis are treated
as evidence against the null. For regression diagnostics, the null
hypothesis would be the assumed model, and visual discoveries could be
any visual features in favour of any alternatives.

In visual inference, a collection of test statistics
\(T^{(i)}(\boldsymbol{\mathrm{y}})~(i \in I)\) is defined, where
\(\boldsymbol{\mathrm{y}}\) is the data and \(I\) is a set of all
possible visual features. \citet{buja_statistical_2009} described each
of the test statistics \(T^{(i)}(\boldsymbol{\mathrm{y}})\) as a
measurement of the degree of presence of a visual feature.
Alternatively, \citet{majumder_validation_2013} avoids the use of visual
features and defined the visual statistics \(T(.)\) as a mapping from a
dataset to a data plot. Both definitions of visual test statistics are
valid, but in the rest of the report the first definition will be used
as it covers some details needed by the following discussion. A visual
discovery is defined as a rejection of a null hypothesis, and the same
null hypothesis can be rejected by many different visual discoveries
\citep{buja_statistical_2009}. For regression diagnostics, the null
hypothesis would be the assumed model, while the visual discoveries
would be any findings that are inconsistent with the null hypothesis.
The same regression model can be rejected by many reasons with residual
plot, including non-linearity and heteroskedasticity as shown in Figure
@ref(fig:residual-plot-cubic-heter).

\begin{figure}
\centering
\includegraphics{paper_comparison_files/figure-latex/residual-plot-cubic-heter-1.pdf}
\caption{Residuals vs.~fitted values plot for a classical linear
regression model. The residuals are produced by fitting a two-predictor
multiple linear regression model with data generated from a cubic linear
model. From the residual plot, ``butterfly shape'' can be observed which
generally would be interpretd as evidence of heteroskedasticity.
Further, from the outline of the shape, nonlinear patterns exist. Both
visual discoveries are evidence against the null hypothesis, though
heteroskedasticity actually does not exist in the data generating
process.}
\end{figure}

\hypertarget{se:sampling-from-null}{%
\subsubsection{Sampling from the null
distribution}\label{se:sampling-from-null}}

The null distribution of plots refers to the infinite collection of
plots of null datasets sampled from \(H_0\). It is defined as the
analogue of the null distribution of test statistics in conventional
test \citep{buja_statistical_2009}. In practice, a finite number of
plots of null datasets could be generated, called null plots. In the
context of regression diagnostics, sampling data from \(H_0\) is
equivalent to sampling data from the assumed model. As
\citet{buja_statistical_2009} suggested, \(H_0\) is usually composited
by a collection of distributions controlled by nuisance parameters.
Since regression models can have various forms, there is no general
solution to this problem, but it sometimes can be reduced to so called
``reference distribution'' by applying one of the three methods: (i)
sampling from a conditional distribution given a minimal sufficient
statistic under \(H_0\), (ii) parametric bootstrap sampling with
nuisance parameters estimated under \(H_0\), and (iii) Bayesian
posterior predictive sampling.

The conditional distribution given a minimal sufficient statistic is the
best justified reference distribution among the three
\citep{buja_statistical_2009}. Suppose there exists a minimal sufficient
statistic \(\boldsymbol{S}(\boldsymbol{y})\) under the null hypothesis,
any null datasets \(\boldsymbol{y^{*}}\) should fulfil the condition
\(\boldsymbol{S}(\boldsymbol{y}) = \boldsymbol{s}\). Using the classical
normal linear regression model as example, the minimal sufficient
statistic is
\(\boldsymbol{S}(\boldsymbol{y}) = (\hat{\boldsymbol{\beta}}, \boldsymbol{e}'\boldsymbol{e})\),
where \(\hat{\boldsymbol{\beta}}\) are the coefficient estimators and
\(\boldsymbol{e}'\boldsymbol{e}\) is the residual sum of square.
Alternatively, the minimal sufficient statistic can be constructed as
\(\boldsymbol{S}(\boldsymbol{y}) = (\hat{\boldsymbol{y}}, ||\boldsymbol{e}||)\),
where \(\hat{\boldsymbol{y}}\) are the fitted values and
\(||\boldsymbol{e}||\) is the length of residuals, which is more
intuitive as suggested by \citet{buja_statistical_2009}. Since the
fitted values are held fixed, the variation can only occur in the
residual space. And because the length of residual is also held fixed,
residuals obtained from a null dataset has to be a random rotation of
\(\boldsymbol{e}\) in the residual space. With this property, null
residuals can be simulated by regressing \(N\) i.i.d standard normal
random draws on the regressors, then rescaling it by the ratio of
residual sum of square in two regressions.

\hypertarget{se:lineup}{%
\subsubsection{Lineup protocol}\label{se:lineup}}

With the simulation of null plots being provided, another aspect of
hypothesis testing that needs to be addressed is the control of false
positive rate or Type I error. Any visual statistic
\(T^{(i)}(\boldsymbol{\mathrm{y}})\) needs to pair with a critical value
\(c^{(i)}\) to form a hypothesis test. When a visual feature \(i\) is
discovered by the observer from a plot, the corresponding visual
statistic \(T^{(i)}(\boldsymbol{\mathrm{y}})\) may not be known as there
is no general agreement on the measurement of the degree of presence of
a visual feature. It is only the event that
\(T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}\) is confirmed. Similarly,
if any visual discovery is found by the observer, we say, there exists
\(i \in I:~T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}\)
\citep{buja_statistical_2009}.

Using the above definition, the family-wise Type I error can be
controlled if one can provide the collection of critical values
\(c^{(i)}~(i \in I)\) such that
\(P(\mathrm{there~exists~} i \in I: T^{(i)}(\boldsymbol{\mathrm{y}}) > c^{(i)}|\boldsymbol{\mathrm{y}}) \leq \alpha\),
where \(\alpha\) is the significance level. However, since the quantity
of \(T^{(i)}(\boldsymbol{\mathrm{y}})\) may not be known, such
collection of critical values can not be provided.

\citet{buja_statistical_2009} proposed the lineup protocol as a visual
test to calibrate the Type I error issue without the specification of
\(c^{(i)}~(i \in I)\). It is inspired by the ``police lineup'' or
``identity parade'' which is the act of asking the eyewitness to
identify criminal suspect from a group of irrelevant people. The
protocol consists of \(m\) randomly placed plots, where one plot is the
actual data plot, and the remaining \(m - 1\) plots have the identical
graphical production as the data plot except the data has been replaced
with data consistent with the null hypothesis. Then, an observer who
have not seen the actual data plot will be asked to point out the most
different plot from the lineup.

Under the null hypothesis, it is expected that the actual data plot
would have no distinguishable difference with the null plots, and the
probability of the observer correctly picks the actual data plot is
\(1/m\). If we reject the null hypothesis as the observer correctly
picks the actual data plot, then the Type I error of this test is
\(1/m\).

This provides us with an mechanism to control the Type I error, because
\(m\) - the number of plots in a lineup can be chosen. A larger value of
\(m\) will result in a smaller Type I error, but the limit to the value
of \(m\) depends on the number of plots a human is willing to view
\citep{buja_statistical_2009}. Typically, \(m\) will be set to \(20\)
which is equivalent to set \(\alpha = 0.05\), a general choice of
significance level for conventional testing among statisticians.

Further, if we involve \(K\) independent observers in a visual test, and
let \(X\) be a random variable denoting the number of observers
correctly picking the actual data plot. Then, under the null hypothesis
\(X \sim \mathrm{Binom}_{K,1/m}\), and therefore, the \(p\)-value of a
lineup of size \(m\) evaluated by \(K\) observer is given as
\begin{equation} \label{eq:pvaluesingle}
P(X \geq x) = \sum_{i=x}^{K}{{K}\choose{i}}\left(\frac{1}{m}\right)^i\left(\frac{m-1}{m}\right)^{k-i},
\end{equation}

where \(x\) is the realization of number of observers correctly picking
the actual data plot \citep{majumder_validation_2013}.

The multiple individuals approach avoids the limit of \(m\), while
provides visual tests with \(p\)-value much smaller than \(0.05\). In
fact, the lower bound of \(p\)-value decreases exponentially as \(K\)
increases. With just \(4\) individuals and \(20\) data plots in a
lineup, the \(p\)-value could be as small as \(0.0001\). Additionally,
by involving multiple observers, variation of individual ability to read
plots can be addressed to some degree as different opinions about visual
discoveries can be collected.

Compared to the conventional test, whose power only depends on the
parameter of interest \(\theta\), several studies
\citep[see][]{hofmann_graphical_2012, majumder_validation_2013, majumder_human_2014, roy_chowdhury_using_2015, loy_variations_2016}
have shown the power of the visual test is subject-specific. Thus, to be
able to account for individual's ability, an individual is required to
evaluate multiple lineups \citep{majumder_validation_2013}.

Suppose individuals have the same ability and a lineup has been
evaluated by multiple individuals, under the alternative hypothesis, the
estimated power for a lineup can be expressed as \(\hat{p} = x/K\), the
estimated probability of identifying the actual data plot from the
lineup. If the individual skill needs to be taken into account, and
\(L\) lineups have been evaluated by \(K\) individuals,
\citet{majumder_validation_2013} suggests that mixed effects logistic
regression model can be fit as:

\[g(p_{li}) = W_{li}\delta + Z_{li}\tau_{li},\] where \(g(.)\) is the
logit link function \(g(p) = log(p) - log(1-p)\); \(0 \leq p \leq 1\).
\(W_{li}\), \(1 \leq i \leq K\), \(1 \leq l \leq L\), is the covariate
matrix including lineup-specific elements and demographic information of
individuals, and \(\delta\) is a vector of parameters. \(Z\) is the
random effects matrix, and \(\tau\) is a vector of variables follow
\(N(\boldsymbol{0},\sigma_{\tau}\boldsymbol{I}_{KL\times KL})\).

Then, the estimated power for lineup \(l\) and individual \(i\) can be
calculated as
\(\hat{p}_{li} = g^{-1}(W_{li}\hat{\delta} + Z_{li}\hat{\tau}_{li})\)
\citep{majumder_validation_2013}.
